---
title: "Mesolithic of Northern Iberia"
subtitle: ""
author: C Michael Barton and Geoffrey A Clark,
date: 'Last Updated: `r Sys.Date()`'
mainfont: Times New Roman
monofont: Monaco
geometry: margin=1in
fontsize: 12pt
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: true
  word_document:
    toc: yes
    toc_depth: '3'
  html_notebook: null
editor_options: 
  chunk_output_type: inline
---
**Copyright (c) C. Michael Barton and Geoffrey A Clark, Arizona State Unversity, 2021. All rights reserved. This software and associated data are licensed under the MIT open source license.** 

```{r setup, warning=FALSE, message=FALSE, results="hide", echo=FALSE}
library(tidyverse)
library(Bchron)
library(rcarbon)
library(doParallel)@
library(parallel)

## Read in raw data file
NIberia_all <- read_csv("NIberia_all.csv",
                        col_types = cols(PG.ID = col_integer(),
                                         C14mean = col_integer(),
                                         C14sdev = col_integer(),
                                         calBPmedian = col_integer(), 
                                         notes = col_character()))

ncores <- detectCores()
options(Ncpus = ncores)
```

# Calibration and SPD with the BCHRON package

## Calibrate dates

### All dates
```{r calibrate}
all.dates.calibrated <- with(NIberia_all, BchronCalibrate(ages = C14mean, ageSds = C14sdev, calCurves = calib.curve, positions = site))

NIberia_all$calBPmedian <- sapply(1:length(all.dates.calibrated), function(x) round(median(all.dates.calibrated[[x]]$ageGrid)))

```


### Filter dates to remove those with high COV
```{r filter}
NIberia_filtered <- NIberia_all %>% 
  mutate(C14.COV=C14sdev/C14mean) %>% 
  dplyr::filter(C14.COV<=0.1)
```


# Calibration, SPD, and modeling with RCARBON package
Dates are binned into 100 year intervals by site

### Compare Cantabria SPD to model

#### Graph results (Figure 18)
``` {r fig18_graph, echo=FALSE, warning=FALSE, message=FALSE, results="hide", fig.width=9, fig.height=5}
date.bins <- NIberia_filtered %>% 
  dplyr::filter(region=="Cantabria", calib.curve != "normal") %>% 
  with(., binPrep(site, C14mean, 100))

model.c <- NIberia_filtered %>% 
  dplyr::filter(region=="Cantabria" & calib.curve != "normal") %>%  
  with(., calibrate(x=C14mean, errors=C14sdev, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  modelTest(., 
            errors = subset(NIberia_filtered, region=="Cantabria" &  calib.curve != "normal")$C14sdev, 
            timeRange = c(13000,4500), 
            runm = 500, 
            model="exponential", 
            datenormalised=TRUE, 
            nsim = 200, 
            ncores = ncores,
            method = 'calsample', 
            bins = date.bins)
  
par(mar=c(7,7,7,3))
plot(model.c, xlim = c(12000,5000), col.obs = 'red', lwd.obs = 5, drawaxes = F)
axis(1, cex.axis = 2, pos = -.004)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=1, line=5, "calibrated years BP", cex=2.5)
mtext(side=2, line=4, "summed probability density", cex=2.5)
title(main=paste("Cantabria Assemblages (N = ", nrow(subset(NIberia_filtered, region=="Cantabria" &  calib.curve != "normal")), ")\n"), cex.main = 3)
```

#### Summarize results
```{r fig18_summary, warning=FALSE, message=FALSE, echo=FALSE}
summary(model.c)
```

*****

### Compare País Vasco SPD to model

#### Graph results (Figure 19)
```{r fig19_graph, warning=FALSE, message=FALSE, results="hide", fig.width=9, fig.height=5}
date.bins <- NIberia_filtered %>% 
  dplyr::filter(region=="País Vasco" & calib.curve != "normal") %>% 
  with(., binPrep(site, C14mean, 100))

model.pv <- NIberia_filtered %>% 
  dplyr::filter(region=="País Vasco", calib.curve != "normal") %>%  
  with(., calibrate(x=C14mean, errors=C14sdev, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  modelTest(., 
            errors = subset(NIberia_filtered, region=="País Vasco" &  calib.curve != "normal")$C14sdev, 
            timeRange = c(13000,4500), 
            runm = 500, 
            model="exponential", 
            datenormalised=TRUE, 
            nsim = 200, 
            ncores = ncores,
            method = 'calsample', 
            bins = date.bins)
 
par(mar=c(7,7,7,3))
plot(model.pv, xlim = c(12000,5000), col.obs = 'red', lwd.obs = 5, drawaxes = F)
axis(1, cex.axis = 2, pos = -.001)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=1, line=5, "calibrated years BP", cex=2.5)
mtext(side=2, line=4, "summed probability density", cex=2.5)
title(main=paste("País Vasco Assemblages (N = ", nrow(subset(NIberia_filtered, region=="País Vasco" &  calib.curve != "normal")), ")"), cex.main = 3)
```
#### Summarize results
```{r fig19_summary, warning=FALSE, message=FALSE, echo=FALSE}
summary(model.pv)
```

*****

### Compare Middle Ebro SPD to model

#### Graph results (Figure 20)
```{r fig20_graph, warning=FALSE, message=FALSE, results="hide", fig.width=9, fig.height=5}
date.bins <- NIberia_filtered %>% 
  dplyr::filter(region=="Ebro", calib.curve != "normal") %>% 
  with(., binPrep(site, C14mean, 100))

model.e <- NIberia_filtered %>% 
  dplyr::filter(region=="Ebro" & calib.curve != "normal") %>%  
  with(., calibrate(x=C14mean, errors=C14sdev, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  modelTest(., 
            errors = subset(NIberia_filtered, region=="Ebro" &  calib.curve != "normal")$C14sdev, 
            timeRange = c(13000,4500), 
            runm = 500, 
            model="exponential", 
            datenormalised=TRUE, 
            nsim = 200, 
            ncores = ncores,
            method = 'calsample', 
            bins = date.bins)
  
par(mar=c(7,7,7,3))
plot(model.e, xlim = c(12000,5000), col.obs = 'red', lwd.obs = 5, drawaxes = F)
axis(1, cex.axis = 2, pos = -.004)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=1, line=5, "calibrated years BP", cex=2.5)
mtext(side=2, line=4, "summed probability density", cex=2.5)
title(main=paste("Middle Ebro Assemblages (N = ", nrow(subset(NIberia_filtered, region=="Ebro" &  calib.curve != "normal")), ")"), cex.main = 3)
```
#### Summarize results
```{r fig20_summary, warning=FALSE, message=FALSE, echo=FALSE}
summary(model.e)
```

*****

### Compare Galicia SPD to model

#### Graph results (Figure 21)
```{r fig21_graph, fig.height=5, fig.width=9, message=FALSE, warning=FALSE, results="hide"}
date.bins <- NIberia_filtered %>% 
  dplyr::filter(region=="Galicia" & calib.curve != "normal") %>% 
  with(., binPrep(site, C14mean, 100))

model.g <- NIberia_filtered %>% 
  dplyr::filter(region=="Galicia", calib.curve != "normal") %>%  
  with(., calibrate(x=C14mean, errors=C14sdev, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>% 
  modelTest(., 
            errors = subset(NIberia_filtered, region=="Galicia" &  calib.curve != "normal")$C14sdev, 
            timeRange = c(13000,4500), 
            runm = 500, 
            model="exponential", 
            datenormalised=TRUE, 
            nsim = 200, 
            ncores = ncores,
            method = 'calsample', 
            bins = date.bins)
  
par(mar=c(7,7,7,3))
plot(model.g, xlim = c(12000,5000), col.obs = 'red', lwd.obs = 5, drawaxes = F)
axis(1, cex.axis = 2, pos = -.0007)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=1, line=5, "calibrated years BP", cex=2.5)
mtext(side=2, line=4, "summed probability density", cex=2.5)
title(main=paste("Galicia Assemblages (N = ", nrow(subset(NIberia_filtered, region=="Galicia" &  calib.curve != "normal")), ")"), cex.main = 3)
```

#### Summarize results
```{r fig21_summary, warning=FALSE, message=FALSE, echo=FALSE}
summary(model.g)
```  

*****
*****

## Permutation Test: comparison of each period with model calculated from all other periods

```{r fig22_graph, echo=FALSE, warning=FALSE, message=FALSE, results="hide"}
date.bins <- NIberia_filtered %>% 
  with(., binPrep(site, C14mean, 100))


alldates_permtest <- NIberia_filtered %>% 
  with(., calibrate(x=C14mean, errors=C14sdev, calCurves = calib.curve, normalised=TRUE, calMatrix=FALSE)) %>%  
  permTest(., marks = NIberia_filtered$region, 
                   timeRange = c(13000,4500),
                   runm = 500, 
                   bins = date.bins, 
                   nsim = 200,
                   datenormalised = TRUE,
                   spdnormalised = FALSE)
```

#### Graph results (Figure 22)

```{r fig22_summary, echo=FALSE, warning=FALSE, message=FALSE, results="hide", fig.width=6, fig.height=10}
par(mfrow=c(4,1))
par(mar=c(7,7,7,3))
par(bty = 'n') 

plot(alldates_permtest,focalm = 1, lwd.obs = 5, col.obs = 'red', drawaxes = F,  xlim = c(12000,5000), ylim = c(0,0.07))
axis(1, cex.axis = 2, pos = -.004)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=2, line=4, "summed probability density", cex=1.5)
title(main=paste("Cantabria (N = ", nrow(subset(NIberia_filtered, region=="Cantabria")), ")"), cex.main = 2.5)

plot(alldates_permtest,focalm = 2, lwd.obs = 5, col.obs = 'red', drawaxes = F,  xlim = c(12000,5000), ylim = c(0,0.11))
axis(1, cex.axis = 2, pos = -.005)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=2, line=4, "summed probability density", cex=1.5)
title(main=paste("Middle Ebro (N = ", nrow(subset(NIberia_filtered, region=="Ebro")), ")"), cex.main = 2.5)

plot(alldates_permtest,focalm = 4, lwd.obs = 5, col.obs = 'red', drawaxes = F,  xlim = c(12000,5000))
axis(1, cex.axis = 2, pos = -.002)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=2, line=4, "summed probability density", cex=1.5)
title(main=paste("País Vasco (N = ", nrow(subset(NIberia_filtered, region=="País Vasco")), ")"), cex.main = 2.5)

plot(alldates_permtest,focalm = 3, lwd.obs = 5, col.obs = 'red', drawaxes = F,  xlim = c(12000,5000))
axis(1, cex.axis = 2, pos = -.001)
axis(2, cex.axis = 2, pos = 12050)
mtext(side=1, line=5, "calibrated years BP", cex=1.5)
mtext(side=2, line=4, "summed probability density", cex=1.5)
title(main=paste("Galicia (N = ", nrow(subset(NIberia_filtered, region=="Galicia")), ")"), cex.main = 2.5)

```

#### Summarize results

```{r, warning=FALSE, message=FALSE, echo=FALSE}
summary(alldates_permtest)
```



```{r echo=FALSE, eval=FALSE}
##Optional to save out data files
save(NIberia_filtered, file = "NIberia_filtered.RDA")
write_csv(NIberia_all, "NIberia_all.csv")
```

